\Huge{Vipassana for Hackers} \\ \large{Paper Two: The Brain} \\ \large\textit{Version 0.2}

### 2019-10-01 paper-two/paper-two.tex
editing round 2: finished DMN

\begin{abstract}
  Two years ago, I took my first Vipassana course and immediately began researching 
the effects of meditation on the brain. Written below is a summary of the research 
available as well as speculations on what goes on in the brain during a 10-day silent 
Vipassana meditation retreat.
\end{abstract}

@@

I took my first Vipassana course (as taught by S.N. Goenka) in 2017. 
Vipassana meditation was so unlike anything I had ever experienced before, 
after the course I was left extremely curious about what exactly had happened 
to me during those ten days. For months afterward, I spent my mornings and 
evenings wading through the research in the field of meditation. I was specifically 
focused on white papers dealing with the effects of a 10-day Vipassana course on 
the brains of participants. The research in this area is limited. 
The quality research which does exist usually uses a sample of highly 
experienced meditators \cite{tibeteanmonks} rather than beginners and/or self-reports rather than
objective measures.

@@

This paper does not contradict Mr. Goenka's sentiment, but instead acts as a starting
point for readers who view the function of the brain as central to the activity of
the mind.

\textbf{Disclaimer:} Although my primary field of research is within the field of
biology, my work is far removed from neuroscience. I have tried to simplify the research available so as to better
understand it. If there are any corrections to make or editing in terms of the
content, please feel free to get in touch with me.

@@

Before dissecting the experience of meditation as it pertains to the brain, this
section describes the different parts of the brain that have shown up in
scientific literature as correlated to meditative practices. Recent neuroscience divides the
brain into the reified geography of the brain, its anatomy, and the abstract concepts
governed by brain activity, its function. Between concrete physiology and abstract
functional outcomes exist networks of cooperative structures which correspond to general
high-level activities of the brain.

The Default Mode Network (DMN) is the constellation of regions which fire when
the brain is not engaged in any external or goal-oriented
tasks. As the name suggests, the brain is in its ``default
mode''. \cite{defaultnetworkadaptive} The DMN is anticorrelated to the Central Executive
Network (CEN), which is active during externally-directed, high-level cognitive
functions. \cite{saliencenetwork} This anticorrelation between the DMN and the CEN is
governed by the Salience Network (SN), the collection of regions in the brain which
help decide which stimuli deserve our attention. The SN acts as a switch between the
internally-directed DMN and the externally-directed CEN. \cite{saliencenetwork}

@@

This network is mostly observed through changes in blood flow to different parts of 
the brain measured using fMRI (functional Magnetic Resonance Imaging) or 
PET (Positron Emission Tomography){defaultnetworkadaptive}

@@

The regions involved in an individual's conceptual` `self'' are the Posterior
Cingulate Cortex (PCC), Ventromedial Prefrontal Cortex (vmPFC), and the Inferior
Parietal Lobule (IPL).That is to say, these regions are preferentially activated when 
volunteers were asked to reflect on personal preferences, beliefs, abilities, 
anything relevant to the self. \cite{defaultnetworkadaptive}

Cerebral blood flow and metabolic rate is 40\% higher than average in the PCC,
making it one of the most active regions of the brain. \cite{pccrole}

The PCC has been thought to mediate interactions between emotion and memory. 
Under an fMRI, it consistently lights up when a person recollects something from their
life. The strength of this activity varies depending on the emotional importance of 
the memory recalled. It is also activated by emotional stimuli, 
both positive and negative, and acts as an interface between the external world 
and the individual by gauging the importance of different stimuli. \cite{pccemotion}

@@

It is hypothesized that the activation of these three regions of the brain (PCC,
vmPFC, and IPL) are responsible for providing a sense of self --- a subjective agent
in space and time. It functions as a network in which phenomena such as self
representations are accessible via the PCC \cite{mappingself} and are gated into 
conscious awareness by activity in the mPFC, as influenced by changing internal 
and external demands. The PCC acts as a brain-wide connectivity hub, through which a unitary notion of self is
created by a large scale integration of DMN activity. \cite{mappingself}

@@

There is a rich body of work pointing to this network (especially the dmPFC subsystem) 
being engaged when individuals reflect upon, evaluate, appraise their own thoughts, 
feelings and desires as well as those of others. \cite{theoryofmind}

In addition to this, the dmPFC is involved in empathy, moral reasoning,
and altruistic behavior. \cite{defaultnetworkadaptive,dmpfcothers,dmpfcaltruism}

@@

The Medial Temporal Subsytem is primarily activated during these processes
This subsystem consists of the Medial Temporal Lobe and parts of the limbic system
especially the Hippocampus.

The limbic system is a complex set of structures that lies on both sides of the
Thalamus, just under the Cerebrum. It includes the Hypothalamus, the Hippocampus, the
Amygdala, and several other nearby areas. It appears to be primarily responsible for
our emotional life, and has a lot to do with the formation of memories. \cite{limbic}

The critical role of the Medial Temporal Lobe (MTL) which is a part of the DMN 
in long term memory is well documented. \cite{mtl}
This network has also been activated during autobiographical memory retrieval, 
contextual association, and semantic knowledge. \cite{defaultnetworkadaptive}

@@

\todo{Summarize DMN? Maybe introduce why it's relevant to meditation?}
Therefore the DMN that plays a role in constructing a sense of self, in memory, 
in thinking about other also lends itself to excessive rumination and anxiety. \cite{dmndepression} 
When I sit down to meditate and my thoughts wander away (and how they wander!),
it is the DMN that is being activated.

@@

There have been studies looking at both the structural changes in the brain and
functional changes in brain activity, especially to the DMN and the SN. But such
studies are usually only done on meditators who have clocked thousands of hours of
practice, which diverges from the purpose of this paper. \cite {tibeteanmonks
, experiencemed, experiencemeda}

@@

\bibitem{tibeteanmonks}
  Alex Hankey
  \url:https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1697747/
  \textit{Studies of Advanced Stages of Meditation in the Tibetan 
           Buddhist and Vedic Traditions. I: A Comparison of General Changes}
   Evid Based Complement Alternat Med. 2006 Dec; 3(4): 513–521.

@@

\bibitem{theoryofmind}
  Frank Van Overwalle
 \textit{A dissociation between social mentalizing and general reasoning}
  NeuroImage. Volume 54, Issue 2, 15 January 2011, Pages 1589-1599
  \url:https://www.sciencedirect.com/science/article/abs/pii/S1053811910012243

\bibitem{limbic}
  Catani M, Dell'Acqua F, Thiebaut De Schotten M
  \textit {A revised limbic system model for memory, emotion and behaviour}
  Neuroscience and Biobehavioral Reviews. 37 (8): 1724–37. 
  doi:10.1016/j.neubiorev.2013.07.001. PMID 23850593.

@@

\bibitem{experiencemed}
   Brewer JA, Worhunsky PD, Gray JR, Tang YY, Weber J, Kober H.
   \textit {Meditation experience is associated with differences in default mode 
          network activity and connectivity.}
     Proc Natl Acad Sci U S A. 2011 Dec 13;108(50):20254-9. 
     doi: 10.1073/pnas.1112029108.

>>>>

\begin{abstract}
  Two years ago, immediately following my first Vipassana course, I began researching
the effects of meditation on the brain. This paper is a summary of the research
available on the topic and speculations, based on my experience, as to what is
happening in the brain during a 10-day silent Vipassana meditation retreat.
\end{abstract}

@@

I took my first Vipassana course (as taught by S.N. Goenka) in 2017.
Vipassana meditation was so unlike anything I had ever experienced before
I was left extremely curious about what exactly had happened
to me during those ten days. For months afterward I spent my mornings and
evenings wading through the research in the field of meditation. I was specifically
focused on white papers dealing with the effects of a 10-day Vipassana course on
the brains of participants. The research in this area is limited.
The quality research which does exist usually uses a sample of highly
experienced meditators \cite{tibetanmonks} rather than beginners and/or self-reports rather than
objective measures.

@@

This paper does not contradict Mr. Goenka's sentiment, but instead acts as a starting
point for curious readers who, whether they have taken a Vipassana course or not,
view the function of the brain as central to the activity of the mind.

\textbf{Disclaimer:} Although my primary field of research is within the field of
biology, my work is far removed from neuroscience. I have tried to simplify the research available so as to better
understand it. If there are any corrections to make or editing in terms of the
content, please feel free to contact me.

@@

Before dissecting the experience of meditation as it pertains to the brain, this
section describes the different parts of the brain that have shown up in
scientific literature as correlated to meditative practices. Recent neuroscience divides the
brain into the reified geography of the brain (anatomy) and the abstract concepts
governed by brain activity (function). Between concrete physiology and abstract
functional outcomes exist networks of cooperative structures which correspond to general
high-level activities of the brain.

The Default Mode Network (DMN) is the constellation of regions which fire when
the brain is not engaged in any external or goal-oriented
tasks. As the name suggests, the brain is in its ``default
mode'' of operation. \cite{defaultnetworkadaptive} The DMN is anticorrelated to the Central Executive
Network (CEN), which is active during externally-directed, high-level cognitive
functions. \cite{saliencenetwork} This anticorrelation between the DMN and the CEN is
governed by the Salience Network (SN), the collection of regions in the brain which
help decide which stimuli deserve our attention. In this way, the SN acts as a switch between the
internally-directed DMN and the externally-directed CEN. \cite{saliencenetwork}

\todo{Introduce both the Limbic System and Somatosensory Cortex here, at a very high level.}

@@

This network is mostly observed through changes in blood flow to different parts of
the brain measured using fMRI (functional Magnetic Resonance Imaging) or
PET (Positron Emission Tomography). \cite{defaultnetworkadaptive}

@@

The regions involved in an individual's conceptual ``self'' are the Posterior
Cingulate Cortex (PCC), Ventromedial Prefrontal Cortex (vmPFC), and the Inferior
Parietal Lobule (IPL). Observationally, these regions are preferentially activated when
volunteers were asked to reflect on personal preferences, beliefs, abilities, or
any other qualities relevant to the self. \cite{defaultnetworkadaptive} Cerebral
blood flow and metabolic rate are 40\% higher than average in the PCC,
making it one of the most active regions of the brain. \cite{pccrole}

The PCC has been thought to mediate interactions between emotion and memory.
Under an fMRI, it consistently lights up when a person recollects an event from their
life. The strength of this activity varies depending on the emotional importance of
the memory recalled. It is also activated by emotional stimuli,
both positive and negative, and acts as an navigational interface, between the individual
and the external world, by gauging the importance of different stimuli. \cite{pccemotion}

@@

\todo{Expand on IPL: Why is facial recognition interesting to this paper?}

It is hypothesized that the activation of these three regions of the brain (PCC,
vmPFC, and IPL) are responsible for providing a sense of self --- a subjective agent
in space and time. It functions as a network in which phenomena such as self
representations are accessible via the PCC \cite{mappingself} and are gated into
conscious awareness by activity in the mPFC \todo{The mPFC needs to be mentioned
  earlier, defined, and abbreviated.} , as influenced by changing internal
and external demands. The PCC acts as a brain-wide connectivity hub, through which a unitary notion of self is
created by a large scale integration of DMN activity. \cite{mappingself}

@@

The DMN (especially the dmPFC subsystem) is engaged when individuals reflect upon,
evaluate, or appraise their own mental states or the mental states of others. ``Mental
states'' include thoughts, feelings, and desires. \cite{theoryofmind} In addition to
consideration of these states, the dmPFC is involved in empathy, moral reasoning,
and altruistic behavior. \cite{defaultnetworkadaptive,dmpfcothers,dmpfcaltruism}

@@

\todo{Define and describe the Medial Temporal Subsystem: Describe MTL anatomically and MTS relation
  to the DMN, particularly the partial inclusion of the LS. Use the Anatomical
  introduction (immediately following the Introduction) to assist in framing this.}

The Medial Temporal Subsytem is primarily activated during these processes
This subsystem consists of the Medial Temporal Lobe (MTL) and parts of the Limbic
System, particularly the Hippocampus.

The Limbic System is a complex set of structures that lies on both sides of the
Thalamus, just under the Cerebrum. It includes the Hypothalamus, the Hippocampus, the
Amygdala, and several other nearby areas. It appears to be primarily responsible for
our emotional life and has a specific involvement in the formation of memories. \cite{limbic}

The critical role of the MTL in long term memory is well documented. \cite{mtl}
This network \todo{Which network?} has also been activated during autobiographical memory retrieval,
contextual association, and semantic knowledge. \cite{defaultnetworkadaptive}

@@

\subsubsection{Summary}

The DMN plays a role in constructing a sense of self, in memory,
and in thinking about others. Because of this, it is also a key component in excessive
rumination and anxiety. \cite{dmndepression}

When a meditator sits down to meditate and thoughts wander away to past experiences
or anxieties about the future, it is the DMN playing the most active role in this
distraction.

\todo{Expand on DMN relationship to Monkey Mind. Add papers with supporting evidence?}

@@

There have been studies looking at both the structural changes in the brain and
functional changes in brain activity, especially to the DMN and the SN. But such
studies are usually only done on meditators who have clocked thousands of hours of
practice, which diverges from the purpose of this paper. \cite{tibetanmonks,
  experiencemed, experiencemeda}

@@

\bibitem{tibetanmonks}
  Alex Hankey
  \url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1697747/}
  \textit{Studies of Advanced Stages of Meditation in the Tibetan
          Buddhist and Vedic Traditions. I: A Comparison of General Changes}
  Evid Based Complement Alternat Med. 2006 Dec; 3(4): 513–521.

@@

\bibitem{theoryofmind}
  Frank Van Overwalle
  \textit{A dissociation between social mentalizing and general reasoning}
  NeuroImage. Volume 54, Issue 2, 15 January 2011, Pages 1589-1599
  \url{https://www.sciencedirect.com/science/article/abs/pii/S1053811910012243}

\bibitem{limbic}
  Catani M, Dell'Acqua F, Thiebaut De Schotten M
  \textit{A revised limbic system model for memory, emotion and behaviour}
  Neuroscience and Biobehavioral Reviews. 37 (8): 1724–37.
  doi:10.1016/j.neubiorev.2013.07.001. PMID 23850593.

@@

\bibitem{experiencemed}
   Brewer JA, Worhunsky PD, Gray JR, Tang YY, Weber J, Kober H.
   \textit{Meditation experience is associated with differences in default mode
           network activity and connectivity.}
   Proc Natl Acad Sci U S A. 2011 Dec 13;108(50):20254-9.
   doi: 10.1073/pnas.1112029108.
